---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tidyheatmap

<!-- badges: start -->
[![Travis build status](https://travis-ci.org/jbengler/tidyheatmap.svg?branch=master)](https://travis-ci.org/jbengler/tidyheatmap)
<!-- badges: end -->

The goal of `tidyheatmap` is to provide a tidyverse-style interface to the powerful heatmap package [pheatmap](https://github.com/raivokolde/pheatmap) by [\@raivokolde](https://github.com/raivokolde). This enables the convenient generation of complex heatmaps from tidy data.

## Installation

You can install `tidyheatmap` from GitHub with:

```{r eval=FALSE}
# install.packages("devtools")

devtools::install_github("jbengler/tidyheatmap")
```

## Example

Given a tidy data frame of gene expression data like `data_exprs`, you can easily generate a basic heatmap. 

```{r eval=FALSE}
library(tidyheatmap)

data_exprs
#> # A tibble: 800 x 3
#>    external_gene_name sample expression
#>    <chr>              <chr>       <dbl>
#>  1 Bsn                Hin_1        9.59
#>  2 Bsn                Hin_2        9.48
#>  3 Bsn                Hin_3        9.66
# ...

tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression
)
```

## Documentation

https://jbengler.github.io/tidyheatmap/


