---
title: "Get started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{tidyheatmap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
collapse = TRUE,
comment = "#>"
)
```

`tidyheatmap` provides a tidyverse-style interface to the powerful heatmap package [pheatmap](https://github.com/raivokolde/pheatmap) by [\@raivokolde](https://github.com/raivokolde). Although heatmaps can be generated via `ggplot2::geom_tile()`, it is hard to reach the versatility and beauty of a genuine heatmap function like `pheatmap::pheatmap()`.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyheatmap)
```

## Data requirements

`tidy_heatmap()` requires tidy data in long format, see [tidyverse](https://tidyr.tidyverse.org/index.html).

As an example we will use the gene expression data set `data_exprs`. In the tidyverse lingo the columns of a data frame are called *variables*. One of these variables named `expression` contains the numeric values to be color-coded in the heatmap. Furthermore we will use the variables `sample` for heatmap columns and `external_gene_name` for heatmap rows.

```{r}
data_exprs

data_exprs %>% filter(external_gene_name == "Bsn") %>% select(external_gene_name, sample, expression)

glimpse(data_exprs)
```

## Basic usage

The basic layout of the heatmap relies on the parameters `rows`, `columns` and `values`. You can think of them like aesthetics in `ggplot2::ggplot()`, similar to something like `aes(x = columns, y = rows, fill = values)`.


```{r}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression
)
```

## Features

### Data scaling

With the parameter `scale` you can activate data scaling for `"row"` or `"column"`. By default data scaling is turned off `scale = "none"`.

```{r}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             scale = "row"
)
```

### Ordering

Rows and columns in the heatmap will appear in the same order as in the tidy data frame used as input. For example, to order rows and columns alphabetically, just use the `dplyr::arrange()`.

```{r}
data_exprs %>% 
  arrange(external_gene_name, sample) %>% 
  tidy_heatmap(rows = external_gene_name,
               columns = sample,
               values = expression,
               scale = "row"
  )
```

### Color legend

You can customize the number of colors `color_scale_n` and also the minimum and maximum values of the color legend, `color_scale_min` and `color_scale_max`.

```{r}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             scale = "row",
             color_scale_n = 16,
             color_scale_min = -2,
             color_scale_max = 2
)
```

Of course, you can also replace the color legend altogether.

```{r}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             scale = "row",
             color = c("#161523","#1C1E4C","#2D3679","#3B4D98","#49569F","#69549D",
                       "#855097","#9E4F96","#CA5296","#DC5F8E","#E17872","#E58D63",
                       "#EDA962","#F5CA66","#F9E96C","#EEE969")
)
```

### Annotations

Annotations can be added for both `rows` and `columns` via `ann_row` and `ann_col`, respectively. Just specify the corresponding variables in the tidy data frame. If you want more then one variable for annotation just combine them by `c(var1, var2, var3)`.

```{r}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             ann_col = c(sample_type, condition, group),
             ann_row = c(is_immune_gene, direction),
             scale = "row",
             color_scale_n = 16,
             color_scale_min = -2,
             color_scale_max = 2
)
```

### Customize annotations colors

You can provide a list of named vectors to take control over the annotations colors `ann_colors`.

```{r}
ann_colors <- list(
  condition = c(EAE = "#BD79B4", healthy = "#F5CEF2"),
  group = c(Ein = "#C14236", Eip = "#E28946", Hin = "#4978AB", Hip = "#98BB85"),
  sample_type = c(input = "#BDBDBD", IP = "#7D7D7D"),
  direction = c(down = "#5071DC", up = "#C34B6B"),
  is_immune_gene = c(yes = "#B69340", no = "#FFFFFF")
)

tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             ann_col = c(sample_type, condition, group),
             ann_row = c(is_immune_gene, direction),
             scale = "row",
             color_scale_n = 16,
             color_scale_min = -2,
             color_scale_max = 2,
             ann_colors = ann_colors
)
```

### Gaps

Gaps can be added by specifying data frame variables that should be used to generate the gaps. Only one variable can be chosen for `gaps_row` and one for `gaps_col`.

```{r}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             ann_col = c(sample_type, condition, group),
             ann_row = c(is_immune_gene, direction),
             scale = "row",
             color_scale_n = 16,
             color_scale_min = -2,
             color_scale_max = 2,
             ann_colors = ann_colors,
             gaps_row = direction,
             gaps_col = group
)
```

### Fix cell dimensions

You can fix the cell dimensions via the `cellwidth` and `cellheight` parameters.

```{r fig.asp=0.8}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             ann_col = c(sample_type, condition, group),
             ann_row = c(is_immune_gene, direction),
             scale = "row",
             color_scale_n = 16,
             color_scale_min = -2,
             color_scale_max = 2,
             ann_colors = ann_colors,
             gaps_row = direction,
             gaps_col = group,
             cellwidth = 7,
             cellheight = 7
)
```

### Write to file

You can use the parameter `filename` to write the heatmap to file.

```{r fig.asp=0.8, eval=FALSE}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             filename = "my_heatmap.pdf"
)
```

## More features

There are even more features like clustering and dendrograms implemented in `pheatmap::pheatmap()`. All parameters that are not directly evaluated by `tidy_heatmap()` will be automatically passed to `pheatmap::pheatmap()`.

```{r fig.asp=0.9}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             ann_col = c(sample_type, condition, group),
             ann_row = c(is_immune_gene, direction),
             scale = "row",
             color_scale_n = 16,
             color_scale_min = -2,
             color_scale_max = 2,
             ann_colors = ann_colors,
             cellwidth = 7,
             cellheight = 7,
             cluster_rows = TRUE,
             cluster_cols = TRUE
)
```

## What's under the hood?

`tidy_heatmap()` takes the tidy data frame and the additional arguments and does the data wrangling needed to satisfy `pheatmap::pheatmap()`. By setting `return_data = TRUE` you can get a glimpse at the result of the data wrangling.

```{r}
tidy_heatmap(data_exprs,
             rows = external_gene_name,
             columns = sample,
             values = expression,
             ann_col = c(sample_type, condition, group),
             ann_row = c(is_immune_gene, direction),
             gaps_row = direction,
             gaps_col = group,
             return_data = TRUE
)
```

## Credit

All the credit for the heatmap functionality goes to [\@raivokolde](https://github.com/raivokolde), the creator of the brilliant [pheatmap](https://github.com/raivokolde/pheatmap) package. `tidyheatmap` just provides a convenient interface to use [pheatmap](https://github.com/raivokolde/pheatmap) in the `tidyverse` ecosystem.

